# Descriptive statistics

How do you summarize numeric data (continuous or discrete) and present your findings? Here we cover how to analyze data over a single numeric variable, both in isolation (e.g. basic summary statistics) and across discrete categories of another variable (e.g. group mean comparison).

Note that you will need the `dcps` data (`DCPS testing.RData`) to replicate the examples in this chapter.

## Describing one variable

### Summary statistics
A useful first step in analyzing the distribution of scores on a single numeric variable is to calculate the relevant summary statistics. Use the `summary()` function for a quick, general overview. This returns the minimum, mean, and maximum scores, as well as the score at 1st, 2nd (median), and 3rd quartiles.

```{r summary, eval = TRUE}
  summary(dcps)  # for every variable in the data frame

  summary(dcps$ProfLang) # for a specific variable
```

For specific inquiries, use the `summarize()` function and customize your report. For example:

```{r summarise, eval=TRUE}
  dcps %>%   # start by piping in the dataset
    summarize(
      Avg = mean(ProfLang), # calculates the mean
      StdDev = sd(ProfLang),   # standard deviation
      Range = max(ProfLang) - min(ProfLang) 
    )
```

### Graphing the distribution
We typically use a histogram or box plot to visualize the distribution of scores on a numeric variable.

```{r descPlots, eval=TRUE, out.width="70%", fig.align='center'}
# Basic histogram
  hist(dcps$ProfLang)

# Basic boxplot
  boxplot(dcps$ProfLang, horizontal = TRUE)
```

See the chapter on data visualization to learn how to format these graphs appropriately for academic or professional settings.

### Testing hypotheses
A one-sample $t-$test (`t.test()`) compares the observed mean on a numeric variable to a hypothesized mean. The resulting $p$-value indicates the probability of observing the mean in your data from a population defined by the null hypothesis (`mu = `).

For example, evaluate the argument that at least half of DC public school pupils read at or above grade level (i.e. $H_0:~\mu \geq 50$).

```{r t1, eval = TRUE}
  t.test(dcps$ProfLang, mu = 50, alternative = 'less')
```

The test results suggest that it is extremely unlikely ($t=-8.6$, $p<0.001$) that we would observe these data if the majority of DC public school pupils read at or above grade level. We can reject the null hypothesis. 


## Omitting missing values

If some observations have missing data (coded `NA`) on a given variable, summary statistic functions may yield an error. To ignore any missing values in carrying out calculations, include the argument `na.rm=TRUE` in the function.

```{r nas, eval=TRUE}
  a = c(1,3,5,NA,7)

  mean(a) # Operation fails
  
  mean(a, na.rm = TRUE) # Successfully calculates the mean
```


## Group comparisons

### Summary statistics by group
Comparing summary statistics across different groups or categories of cases requires that you identify the grouping variable (typically `group_by()`} before calculating the desired summary statistics.

```{r grpby, eval=TRUE, warning=FALSE, message=FALSE}
  dcps %>%
    group_by(SchType) %>%  # identify grouping variable
    summarize(
      Avg = mean(ProfMath),  # apply functions to each group
      StDev = sd(ProfMath)
    )
```

### Visualize group differences
One good way of exploring these relationships graphically is to use a boxplot. Note that the syntax is to identify the outcome before the `~` and the grouping variable after (`boxplot(OutcomeVar ~ GroupVar, Data)`):

```{r groupbox,eval=TRUE,collapse=TRUE,out.width="70%",fig.align='center'}
  boxplot(ProfMath ~ SchType, data = dcps)
```

### Testing hypotheses
You can test hypotheses about the relationship between a nominal exposure variable (`GroupingVar`) and a numeric outcome (`OutcomeVar`) using either the $t-$ or $F-$ test. Use `t.test(OutcomeVar ~ GroupVar, Data)` to compare the mean outcome across exactly two groups (i.e. when `GroupVar` is binary). Use `aov(OutcomeVar ~ GroupVar, Data)` when `GroupVar` identifies more than two categories. Note that you need to use `summary()` to view the full `aov()` estimates.

Evaluate the argument that large schools (i.e. where more than 200 students took the test) have a significantly different level of math proficiency than small schools. 
```{r grouptest,eval=TRUE}
# t-test (two-group test of equivalence)
  t.test(ProfMath ~ (NumTested > 200), data = dcps)
```

While there is a mean difference in math proficiency (31 vs 25), the difference is not statistically significant ($p=0.257$). 

Next consider the possibility that math proficiency differs systematically across type of school. Because `SchType` has more than two values (Elementary, Middle, and High), we use the $F-$test.

```{r anova, eval=TRUE}
# F-test (multigroup test of equivalence)  
  ftest = aov(ProfMath ~ SchType, data = dcps)
  summary(ftest) # view the results of the F-test
```

Here the results indicate a significant difference in math proficiency by type of school ($F_{2,105}=8.35$, $p<0.001$).
