# Frequency and cross tabulation

Nominal data require a different approach. Rather than summary statistics, the relevant information here is the frequency with which each value or category appears in the data. Here we cover how to analyze data for a single nominal variable (e.g. frequency tables and bar charts) and how to evaluate relationships between nominal variables (cross-tabulation).

Note that you will need the `film` data (`biopics.xls`) to replicate the examples in this chapter.


## Describing one variable

### Frequency tables
The `count()` function in `tidyverse` creates a tibble with each value of the variable and the "count" of observations within. 

```{r rawtab, eval=TRUE}
# Frequency table
  Tab =
    film %>%   # the dataset
    count(SubjectSex) # the variable to count
  
  Tab
```

Calculating the the percent of total cases in each category (relative frequency) requires an extra line of code (`mutate(Percent = 100 * n/sum(n))`).

```{r reltab, eval=TRUE}
# Relative frequency
  Tab = 
    Tab %>%
    mutate(Percent = 100 * n/sum(n))
  
  Tab
```

### Bar charts
Use a basic `barplot()` to display the results saved into these objects (`rawTab`). Note the notation here is `barplot(count ~ category, data)`:

```{r bargraph, eval=TRUE, out.width="70%", fig.align='center'}
  barplot(n ~ SubjectSex, Tab)

  barplot(Percent ~ SubjectSex, Tab)
```


## Cross-tabulation
There is a three-step process for presenting and evaluating the association between two nominal variables. The first step is to create a basic cross-tabulation of the joint frequency using the `count()` function. Here we consider the possibility that the representation of female subjects in biopics (`SubjectSex`) has changed over time (`Period`).

```{r xtabct, eval=TRUE}
# Raw cross-tabulation
  xtab =
    film %>% 
    count(SubjectSex,Period) %>%  # (OutcomeVar,ExposureVar)
    na.omit() %>% # drop NA categories
  # now organize results into a 2-way table
    pivot_wider(
      names_from = Period, # MUST be the ExposureVar
      values_from = n, 
      values_fill = 0
    )

  xtab
```

Second, use `chisq.test()` to conduct a $\chi^2$ test of independence. Specify the contingency table created above and add `[-1]` to exclude the first column (category names) from the calculation/

```{r chi2, eval=TRUE}
  chisq.test(xtab[-1])
```

Based on these results, the given sex of biopic subjects is independent of (ie does not differ systematically across) time period. The relationship is not statistically significant ($\chi^2(2,N=761)=0.40$, $p=0.81$).

Finally, to present the results of a cross-tabulation, convert the raw frequencies in your table to percentages (within categories of the the exposure variable). Start by calling the raw tabulation from above.

```{r xtabrel, eval=TRUE}
# Relative freq for presentations  
  xtab %>%
  # add a row total
    mutate(Total = rowSums(.[-1])) %>%
  # convert to percentage
    mutate_at(-1, ~ round(100 * ./sum(.), digits=1))
```

Copy and paste the table into your document. Then format appropriately (e.g. category labels) for final presentation.

