# Doing more with data frames

Very few data sets come "ready made" for analysis. Even in our introductory course, there are times when you will need to limit your analysis to certain observations, recode a variable, and more. This chapter covers a few basics of data "munging."

Note that you will need the `film` (`"biopics.xls"`) and `dcps` (`"DCPS testing.RData"`) data sets to replicate the commands.

## Filter/subset data
It is often necessary to limit your analysis to some subset of cases. Use the `filter()` command to specify the criteria by which to select cases.

```{r filter,eval=TRUE}
  film =
    film %>%
    filter(SubjectSex == 'Female') # criteria for keeping cases

  head(film)
```

The conditions inside `filter()` identify the cases, or rows, to keep (i.e. you're selecting only those rows that satisfy the given conditions). This can be based on any number of conditions. Note that a double equals sign `==` is used to check a logical condition.


## Selecting your variables
A dataframe with loads of variables can be unwieldy. Use the `select()` function to isolate the variables you need for your analysis.

```{r select,eval=TRUE}
  Study1 =
    film %>%
    select(Title,Release,SubjectSex)

  head(Study1)
```

## Create new variables
The `mutate()` function creates new variables (or columns) in a data frame. By assigning the same object name to the result, the new variables become part of the object.

```{r mutate,eval=TRUE}
  dcps =
    dcps %>%
    mutate(
      LastEdit = '2021',  # character variable with same value in each
      ProfLangProp = ProfLang/100, # convert to proportion
    )
  
  dcps %>% select(2,5,8:9) %>% head()
```

In addition to transformations that apply the same function to every case, you can create variables that treat cases on an individual basis. Consider the `if_else()` option. This will specify a logical test to apply to each case, and specify replacement values for those cases that evaluate as true and those cases that evaluate as false. The syntax is: `if_else(criteria,value.if.true,value.if.false)`.

```{r ifelse,eval=TRUE}
  dcps =
    dcps %>%
    mutate(
      AbvMedMath = if_else(
        ProfMath > median(ProfMath), # condition to evaluate
        1,  # output if condition is TRUE
        0   # output if FALSE
      )
    )

  dcps %>% select(SchCode,ProfMath,AbvMedMath) %>% head()
```

## Appending and merging data

It is often useful to combine data from different sources. This may take the form of appending (adding additional cases with information on the same variables) or merging (adding additional variables that describe the same cases).

### Appending new cases
To append new cases to your data frame, use `bind_rows(OldData,NewData)`. Note that the variable names need to match exactly across data frames, but the variable order does not matter.

```{r append,eval=TRUE}
  # old data
    myData = tribble(
      ~District, ~Students,
      115, 985,
      116, 1132
    )

  # new data to add
    new = tribble(
      ~District, ~Students,
      117, 419,
      118, 633
    )
    
  # Append new to old
    myData = bind_rows(myData,new)
    
    myData
```

### Merging
To merge data frames (add new variables for existing cases), use `left_join(OldData,NewData)`. In order to link rows in one data frame to rows in another, it is critical that the data sets contain a common identifier, with the same variable name and same values. Building on the example above

```{r merge,eval=TRUE}
  # new variables
    newvars = tribble(
      ~District,~Teachers,
      115, 43,
      116, 71,
      118, 55
    )

  # join new to old
    myData = left_join(myData,newvars)
    
    myData
```
